|

### 2023-06-07 manuscript/paper.Rmd
trying again?
updated bibliography and some intro methods text

Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed (reviewed in @Herrmann2021). Niche changes in habitat and resource use should then drive phenotypic evolution as the population adapts to new conditions (@Herrmann2021). For example, resource use expansion should result in trophic trait evolution (cite; e.g.,). Habitat expansion should result in locomotor change (cite; e.g.,). And release from enemies should result in reduction of costly defense traits (cite; e.g.,).

Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use and associated phenotypes (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Such character displacement between the sexes is therefore one explanation for sexual dimorphism (de Lisle papers).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution lineage of the fossil stickleback fish (Gasterosteus doryssus). The sex of the individual fossils, which is unobserved in practice, was imputed using multiple imputation by chained equations (MICE) (@MICE) with observed data used to train the imputation model based on extant stickleback with known sex.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense. The fossil sequence captures a bout of adaptive evolution following ecological release from predators, as follows. The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; Roesti et al. 2023). The lineage thus adaptively evolved armor loss and reduction in several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al. in prep), and tooth wear data suggest that the population also began eating more planktonic prey, expanding to an open water lifestyle (Purnell et al. 2007).

Sexual Dimorphism is interesting because......

Sexual Dimoprhism lit review:
@SaittaEtAl2020

The remainder of this manuscript contains a description of the data in section \ref{sec:data} and a description of our models in \ref{sec:models}. Section \ref{sec:results} presents a summary of our results, and we end with our conclusion and future work in section \ref{sec:conclusions}.

@@

A major challenge in this work is inferring the sex of fossil specimens. This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020). Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).

We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G. doryssus is part of the extant threespine stickleback species complex (Gastersteus aculeatus).  Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.

>>>>

Ecological release theory suggests that a prey population’s niche should shift and expand when predation is relaxed or removed (reviewed in @Herrmann2021). 
Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Accordingly, sexes should also diverge in morphological phenotypes associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (de Lisle papers).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution, 16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population also began eating more planktonic prey, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect to observe sexual dimorphism evolution for traits related to habitat use in <i>G. doryssus</i>.

We imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense to test whether increased sexual dimorphism is associated with ecological release from predators.

@@

A major challenge in this work is inferring the sex of fossil specimens. This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020). Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020). We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G. doryssus is part of the extant threespine stickleback species complex (Gasterosteus aculeatus).  Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.
